---
title: "sim"
author: Report by Rich Evans
date: 2023
output:
    html_document


---

This simulation implements part of:

Gu W, Swihart RK. Absent or undetected? Effects of non-detection of species occurrence on wildlife–habitat models. Biological conservation. 2004 Apr 1;116(2):195-203.

But they are random sampling a wildlife area. They randomly pick occupied areas in the simulation. In other words they have a grid and randomly assign the occupied areas. 

I have retrospective data there. I use a mixture of normals because I have sample sizes I need to investigate

```{r,include=FALSE}
# install.packages("pacman")
pacman::p_load(
  data.table, # to group and clean data
  tidyverse, # allows use of pipe (%>%) function in this chapter
  janitor, # for clean_names and tabyl
  broom, # tidy output, use tidy(xxx)
  gtsummary, # nice tables
  magrittr, # for %<>%
  effectsize,
  pROC,
  rstatix,
  pwr, # power calculations
  pander, # to write tables as simple text for cut-and-paste
  knitr,
  kableExtra
)
```




```{r,include=FALSE}
sim.fun <- function(
    num.iter = NULL,
    n.cases = NULL,
    n.naiv.control = NULL,
    eff.sz = NULL,
    prp.nondet = NULL, ...) {
  out1 <- NULL
  out2 <- NULL
  out3 <- NULL
  out4 <- NULL

  ### numbers of things
  n.dogs <- n.cases + n.naiv.control # total number of dogs


  # control group is made up of the unlabeled pos and the real unaffected
  unlab.pos <- round(n.naiv.control * prp.nondet) # number of positive dogs who are unlabled
  n.tru.unaff <- n.naiv.control - unlab.pos # the number of unaffected

  n.tru.aff <- n.cases + unlab.pos # the number of positive dogs in both groups

  tru.labels <- c(rep(1, n.tru.aff), rep(0, (n.dogs - n.tru.aff)))
  naive.labels <- c(rep(1, n.cases), rep(0, n.naiv.control))



  for (i in 1:num.iter) {
    ### sample X
    # this is where the stocastic part comes in

    x.tru.affected <- rnorm(n.cases, mean = eff.sz)

    x.tru.control <- c(
      rnorm(unlab.pos, mean = eff.sz),
      rnorm(n.tru.unaff)
    )

    naive.dat <- data.frame(
      X = c(x.tru.affected, x.tru.control),
      y = naive.labels
    )

    m.naive <- glm(y ~ X, data = naive.dat, family = binomial())

    naive.coeff <- m.naive$coefficients["X"]
    naive.p <- summary(m.naive)$coef[2, 4] # pvalue

    out1 <- c(out1, naive.p)
    out2 <- c(out2, exp(naive.coeff))
    out3 <- c(out3, unlab.pos)

    # temp.out1 <- cohens_d(data=naive.dat,X ~ y,var.equal=FALSE)[1,4,drop=TRUE]


    temp.out4 <- pwr::pwr.t2n.test(
      n1 = n.cases,
      n2 = n.naiv.control,
      sig.level = 0.05,
      alternative = "two.sided",
      power = NULL,
      d = temp.out1
    )$power
    out4 <- c(out4, temp.out4)
  } # loop

  return(list(
    n.cases = n.cases,
    n.naiv.control = n.naiv.control,
    prp.nondet = prp.nondet,
    eff.sz = eff.sz,
    naive.p = out1,
    naive.or = out2,
    unlabeled.pos = out3,
    naive.pwr = out4
  ))
}
```





```{r,results='asis'}
sim.table <- function(parmlist) {
  now <- Sys.time()

  foo <- pmap(parmlist, sim.fun)


  goo <- bind_rows(foo, .id = "column_label") # unlists and binds rows



  zoo1 <- goo %>%
    group_by(n.cases, n.naiv.control, prp.nondet, eff.sz) %>%
    dplyr::summarise(
      prp.nondet = mean(prp.nondet),
      eff.sz = mean(eff.sz),
      naive.or.mn = mean(naive.or),
      naive.p.mn = mean(naive.p),
      naive.p.sd = sd(naive.p),
      unlabeled.pos = mean(unlabeled.pos),
      naive.power = mean(naive.pwr)
    )

  print(difftime(Sys.time(), now, units = "mins"))
  return(pandoc.table(zoo1 %>% ungroup() %>% arrange(prp.nondet),
    style = "rmarkdown", digits = 3, split.table = Inf
  ))
}


oddsratio_to_d(2, log = FALSE) # this coverts OR to cohens so that I make this like SNPs data
d_to_oddsratio(.1)
oddsratio_to_d(1.2, log = FALSE)
```
<!-- https://cran.r-project.org/web/packages/pander/vignettes/pandoc_table.html -->

```{r, results='asis',message=FALSE,warning=FALSE}
### this checlks different patterns for the same sample size



#  print("sample sizes check")
#  parmlist<-list(
# c(rep(5,6)),             #interations; num.iter
# rep(c(150,50,100),2),   #n.cases
# rep(c(50,150,100),2),   #number unlabeled
# c(rep(.5,6)),  #eff.sz
# c(rep(.1,3),rep(.001,3)) #non-detection proportion
# )
#
# sim.table(parmlist=parmlist)


# print("sample sizes from the papers: healey (216)")
#  parmlist<-list(
# c(rep(7555,21)),             #interations; num.iter
# rep(c(161,150,108,121,100,95,161),3),  #cases; n.cases
# rep(c(55,55,108,95,100,95,50),3),  #controls unlabeled
# c(rep(0.5,21)), #treat effect
# c(rep(0.001,7),rep(.05,7),rep(.1,7)) #non-detection proportion
#  )
# sim.table(parmlist=parmlist)
#
#
#
#
# print("sample sizes from the papers: conz , ")
#  parmlist<-list(
# c(rep(7555,3)),             #interations; num.iter
# rep(191,3),  #cases; n.cases
# rep(143,3),  #controls unlabeled
# c(rep(0.3,3)), #treat effect
# c(.001,.05,.1) #non-detection proportion
#  )
# sim.table(parmlist=parmlist)
#
#
# print("sample sizes from the papers: healey (216),innes, conz , baker")
#  parmlist<-list(
# c(rep(7555,12)),             #interations; num.iter
# rep(c(91,161,191,156),3),  #cases; n.cases
# rep(c(126,55,143,241),3),  #controls unlabeled
# c(rep(0.5,12)), #treat effect
# c(rep(.001,4),rep(.05,4),rep(.1,4)) #non-detection proportion
#  )
# sim.table(parmlist=parmlist)
#
#
#
# print("sample sizes check")
#  parmlist<-list(
# c(rep(755,3)),             #interations; num.iter
# rep(c(150,50,100),1),   #n.cases
# rep(c(50,150,100),1),   #number unlabeled
# c(rep(.5,3)),  #eff.sz
# c(rep(.1,3)) #non-detection proportion
# )

# sim.table(parmlist=parmlist)

# print("sample sizes check")
#  parmlist<-list(
# c(rep(7555,60)),             #interations; num.iter
# rep(c(150,50,100),20),   #n.cases
# rep(c(50,150,100),20),   #number unlabeled
# c(rep(.5,60)),  #eff.sz
# c(sort(rep(seq(.005, .1,by=0.005),3))) #non-detection proportion
# )
#
# sim.table(parmlist=parmlist)
```


This is a stand alone piece that saves the output table so I can view the results different ways.
```{r}
now <- Sys.time()

# parmlist<-list(
# c(rep(200,60)),             #interations; num.iter
# rep(c(150,50,100),20),   #n.cases
# rep(c(50,150,100),20),   #number unlabeled
# c(rep(.5,60)),  #eff.sz
# c(sort(rep(seq(.005, .1,by=0.005),3))) #non-detection proportion
# )
#

# parmlist<-list(
# c(rep(7000,10)),             #interations; num.iter
# rep(c(120,80),5),   #n.cases
# rep(c(80,120),5),   #number unlabeled
# c(rep(.5,10)),  #eff.sz
# c(sort(rep(seq(.005, .1,by=0.02),2))) #non-detection proportion
# )

parmlist <- list(
  c(rep(5000, 11)), # interations; num.iter
  rep(c(100), 11), # n.cases
  rep(c(100), 11), # number unlabeled
  c(rep(.5, 11)), # eff.sz
  c(sort(rep(seq(0, .1, by = 0.01), 1))) # non-detection proportion
)


# parmlist<-list(
# c(rep(5000,8)),             #interations; num.iter
# rep(c(110,100),4),   #n.cases
# rep(c(90,100),4),   #number unlabeled
# c(rep(.5,8)),  #eff.sz
# c(sort(rep(c(0,.01,.05,.1),2))) #non-detection proportion
# )


big.otpt <- bind_rows(pmap(parmlist, sim.fun), .id = "column_label") # unlists and binds rows



smry.otpt <- big.otpt %>%
  group_by(n.cases, n.naiv.control, prp.nondet, eff.sz) %>%
  dplyr::summarise(
    prp.nondet = mean(prp.nondet),
    eff.sz = mean(eff.sz),
    naive.or.mn = mean(naive.or),
    naive.p.mn = mean(naive.p),
    naive.p.sd = sd(naive.p),
    unlabeled.pos = mean(unlabeled.pos),
    naive.power = mean(naive.pwr)
  )


smry.otpt %>%
  ungroup() %>%
  select(-c(naive.or.mn, naive.p.mn, naive.p.sd, eff.sz)) %>%
  arrange(prp.nondet, naive.power)

smry.otpt %>%
  ungroup() %>%
  select(-c(naive.or.mn, naive.p.mn, naive.p.sd, eff.sz)) %>%
  arrange(naive.power)

pandoc.table(smry.otpt %>% ungroup() %>% select(-c(naive.or.mn, naive.p.mn, naive.p.sd, eff.sz)) %>% arrange(naive.power),
  style = "rmarkdown", digits = 3, split.table = Inf
)

print(difftime(Sys.time(), now, units = "mins"))


smry.otpt %>%
  ungroup() %>%
  select(n.cases, n.naiv.control, prp.nondet, naive.power) %>%
  reframe(prp.nondet = prp.nondet, zzz = naive.power / mean(naive.power[prp.nondet == 0]))
```


```{r}
df.plot <- smry.otpt %>%
  ungroup() %>%
  select(n.cases, n.naiv.control, prp.nondet, naive.power) %>%
  reframe(prp.nondet = prp.nondet, zzz = naive.power / mean(naive.power[prp.nondet == 0])) %>%
  mutate(prp.nondet = prp.nondet * 100, zzz = zzz * 100)

tiff("powerloss.tiff", units = "in", width = 6, height = 4, res = 600)

ggplot(data = df.plot, aes(x = prp.nondet, zzz)) +
  geom_line(color = "red") +
  geom_point() +
  xlab("Percent of non-detected positives in the control group") +
  ylab("Relative power (%)") +
  ggtitle("Power loss from positive cases in the negative control group")
dev.off()
```

