---
title: "sim"
author: Report by Rich Evans
date: 2023
output:
    html_document


---

This simulation implements part of:

Gu W, Swihart RK. Absent or undetected? Effects of non-detection of species occurrence on wildlife–habitat models. Biological conservation. 2004 Apr 1;116(2):195-203.

But they are random sampling a wildlife area. They randomly pick occupied areas in the simulation. In other words they have a grid and randomly assign the occupied areas. 

I have retrospective data there.

```{r,include=FALSE}
#install.packages("pacman")
pacman::p_load(
       data.table, # to group and clean data
       tidyverse,  # allows use of pipe (%>%) function in this chapter
       janitor,     #for clean_names and tabyl
       broom,   #tidy output, use tidy(xxx)
       gtsummary,  #nice tables
       magrittr, # for %<>%
       effectsize,
       rstatix,
       pwr,   # power calculations
       pander, #to write tables as simple text for cut-and-paste
       knitr,
        kableExtra
   ) 
```




```{r,include=FALSE}

sim.fun<-function(num.iter = NULL
                  ,n.cases = NULL
                  ,n.naiv.control = NULL
                  ,eff.sz=NULL
                  ,prp.nondet = NULL,...){
out1 <- NULL
out2 <- NULL
out3 <- NULL
out4 <- NULL

n.dogs<-n.cases + n.naiv.control

tru.prev<-(n.cases + (n.naiv.control*prp.nondet))/n.dogs

for (i in 1:num.iter){

### calculate the observed prevalence
# N=217 with 55 controls and 161 cases.
# the true prevalence is (161 + (55*prop.nondect))/217
# if prop.nondetect=0.1 then with rounding we should have 161+6 positives
# six are mislabeled as negative.
#how can I sample so that on average 161+6 subjects are postitive.
#by changing the cutoff for the runif to (161+6)/217=0.77
  

  
###sample X  and transform it
# truth
X <- exp(rnorm(n.dogs))

trans.1<-exp(-4+5*X)

preds<- trans.1/(1+trans.1)

#preds<-runif(216)
### add random detection for affected but not yet misclassified
#need the average of the runif= 0.77
tru.labels<- as.numeric(preds>runif(n.dogs,max=2*(1-tru.prev)))

### estimate coeff before misclassification (tru model)

tru.dat<-data.frame(X=X,
                       y=tru.labels)

m.tru<-glm(y~X,data=tru.dat, family = binomial())
tru.coeff<-m.tru$coefficients['X']

### make random misclassification

naive.labels<- tru.labels

# affected get reassigned to zero if runif<nondet prop, one if > prop. non detected

# a 0-1 vector, affected get reassigned to zero if runif<nondet prop, one if
naive.labels[tru.labels==1]<- as.numeric(runif(sum(tru.labels))>prp.nondet) 

### estimate coeff naive model
naive.dat<-data.frame(X=X,
                       y=naive.labels)
m.naive<-glm(y~X,data=naive.dat, family = binomial())
naive.coeff<-m.naive$coefficients['X']



#rel.bias <- 100*(naive.coeff-tru.coeff)/abs(tru.coeff)

out1<- c(out1,exp(naive.coeff))

out2<- c(out2,sum(tru.labels))
out3<- c(out3,sum(naive.labels))
 
} #loop

return(list(n.cases=n.cases,
            n.naiv.control=n.naiv.control,
                  prp.nondet=prp.nondet,
                  naive.or=out1,
                  truprp=out2,
                  navprp=out3))
}


                    
```





```{r,results='asis'}


sim.table<-function(parmlist){
 
  now<-Sys.time()
  
foo<-pmap(parmlist,sim.fun)


goo <- bind_rows(foo, .id = "column_label")  #unlists and binds rows


zoo1 <- goo %>% group_by(n.cases,n.naiv.control,prp.nondet) %>% 
  dplyr::summarise(prp.nondet=mean(prp.nondet)
                   ,naive.or=mean(naive.or)
                   ,truprp=mean(truprp)
                   ,navprp=mean(navprp))

print(difftime(Sys.time(),now,units="mins"))
return(pandoc.table(zoo1 %>% ungroup() %>% arrange(desc(naive.or))
                    ,style='rmarkdown',digits=3,split.table=Inf))

}


oddsratio_to_d(2,log=FALSE) #this coverts OR to cohens so that I make this like SNPs data
d_to_oddsratio(.1)
oddsratio_to_d(1.2,log=FALSE)

 
```
<!-- https://cran.r-project.org/web/packages/pander/vignettes/pandoc_table.html -->

```{r, results='asis',message=FALSE,warning=FALSE}
###this checlks different patterns for the same sample size



# print("sample sizes from the papers: healey (216)")
#  parmlist<-list(
# c(rep(200,4)),             #interations; num.iter
# c(161,161,161,161),   #n.dogs
# c(55,55,55,55),   #n.dogs
# c(.01,.05,.1,.2) #non-detection proportion
# 
#  )
# 
# sim.table(parmlist=parmlist)


 print("sample sizes from the papers: healey (216)")
 parmlist<-list(
c(rep(500,3)),             #interations; num.iter
c(150,50,100),   #n.cases
c(50,150,100),   #n.control
c(rep(0.1,3)) #non-detection proportion

 )

sim.table(parmlist=parmlist)

```

