---
title: "sim"
author: Report by Rich Evans
date: 2023
output:
    html_document


---

This simulation implements part of:

Gu W, Swihart RK. Absent or undetected? Effects of non-detection of species occurrence on wildlife–habitat models. Biological conservation. 2004 Apr 1;116(2):195-203.

But they are random sampling a wildlife area. They randomly pick occupied areas in the simulation. In other words they have a grid and randomly assign the occupied areas. 

I have retrospective data there. I use a mixture of normals because I have sample sizes I need to investigate

```{r,include=FALSE}
# install.packages("pacman")
pacman::p_load(
    data.table, # to group and clean data
    tidyverse, # allows use of pipe (%>%) function in this chapter
    janitor, # for clean_names and tabyl
    broom, # tidy output, use tidy(xxx)
    gtsummary, # nice tables
    magrittr, # for %<>%
    effectsize,
    pROC,
    rstatix,
    pwr, # power calculations
    pander, # to write tables as simple text for cut-and-paste
    knitr,
    rcompanion, # for cohen's w
    kableExtra
)
```




```{r,include=FALSE}
sim.fun <- function(
    num.iter = NULL,
    n.cases = NULL,
    n.naiv.control = NULL,
    eff.sz = NULL,
    prp.nondet = NULL, ...) {
    out1 <- NULL
    out2 <- NULL
    out3 <- NULL
    out4 <- NULL

    ### numbers of things
    n.dogs <- n.cases + n.naiv.control # total number of dogs


    # control group is made up of the unlabeled pos and the real unaffected
    # number of positive dogs who are unlabled
    unlab.pos <- round(n.naiv.control * prp.nondet)
    n.tru.unaff <- n.naiv.control - unlab.pos # the number of unaffected

    n.tru.aff <- n.cases + unlab.pos # the number of positive dogs in both groups

    tru.labels <- c(rep(1, n.tru.aff), rep(0, (n.dogs - n.tru.aff)))
    naive.labels <- c(rep(1, n.cases), rep(0, n.naiv.control))



    for (i in 1:num.iter) {
        ### sample X
        # this is where the stocastic part comes in

        x.tru.affected <- rnorm(n.cases, mean = eff.sz)

        x.tru.control <- c(
            rnorm(unlab.pos, mean = eff.sz),
            rnorm(n.tru.unaff)
        )

        naive.dat <- data.frame(
            X = as.numeric(c(x.tru.affected, x.tru.control) > 0.5),
            y = naive.labels
        )

        m.naive <- glm(y ~ X, data = naive.dat, family = binomial())

        naive.coeff <- m.naive$coefficients["X"]
        naive.p <- summary(m.naive)$coef[2, 4] # pvalue

        out1 <- c(out1, naive.p)
        out2 <- c(out2, exp(naive.coeff))
        out3 <- c(out3, unlab.pos)

        temp.out1 <- rcompanion::cohenW(naive.dat$y, naive.dat$X) # cohen_w


        # temp.out4 <- pwr::pwr.t2n.test(
        #     n1 = n.cases,
        #     n2 = n.naiv.control,
        #     sig.level = 0.05,
        #     alternative = "two.sided",
        #     power = NULL,
        #     d = temp.out1
        # )$power
        # out4 <- c(out4, temp.out4)

        temp.out4 <- pwr.chisq.test(w = temp.out1, df = 1, N = 200, sig.level = 0.05)$power
        out4 <- c(out4, temp.out4)
    } # loop

    return(list(
        n.cases = n.cases,
        n.naiv.control = n.naiv.control,
        prp.nondet = prp.nondet,
        eff.sz = eff.sz,
        naive.p = out1,
        naive.or = out2,
        unlabeled.pos = out3,
        naive.pwr = out4
    ))
}
```





```{r,results='asis'}
sim.table <- function(parmlist) {
    now <- Sys.time()

    foo <- pmap(parmlist, sim.fun)


    goo <- bind_rows(foo, .id = "column_label") # unlists and binds rows



    zoo1 <- goo %>%
        group_by(n.cases, n.naiv.control, prp.nondet, eff.sz) %>%
        dplyr::summarise(
            prp.nondet = mean(prp.nondet),
            eff.sz = mean(eff.sz),
            naive.or.mn = mean(naive.or),
            naive.p.mn = mean(naive.p),
            naive.p.sd = sd(naive.p),
            unlabeled.pos = mean(unlabeled.pos),
            naive.power = mean(naive.pwr)
        )

    print(difftime(Sys.time(), now, units = "mins"))
    return(pandoc.table(zoo1 %>% ungroup() %>% arrange(prp.nondet),
        style = "rmarkdown", digits = 3, split.table = Inf
    ))
}

# this coverts OR to cohens so that I make this like SNPs data
oddsratio_to_d(2, log = FALSE)
d_to_oddsratio(.1)
oddsratio_to_d(1.2, log = FALSE)
```
<!-- https://cran.r-project.org/web/packages/pander/vignettes/pandoc_table.html -->



This is a stand alone piece that saves the output table so I can view the results different ways.
```{r}
now <- Sys.time()


parmlist <- list(
    c(rep(10000, 11)), # interations; num.iter
    rep(c(150), 11), # n.cases
    rep(c(50), 11), # number unlabeled
    c(rep(.5, 11)), # eff.sz for normals
    c(sort(rep(seq(0, .1, by = 0.01), 1))) # non-detection proportion
)


big.otpt <- bind_rows(pmap(parmlist, sim.fun),
    .id = "column_label"
) # unlists and binds rows



smry.otpt <- big.otpt %>%
    group_by(n.cases, n.naiv.control, prp.nondet, eff.sz) %>%
    dplyr::summarise(
        prp.nondet = mean(prp.nondet),
        eff.sz = mean(eff.sz),
        naive.or.mn = mean(naive.or),
        naive.p.mn = mean(naive.p),
        naive.p.sd = sd(naive.p),
        unlabeled.pos = mean(unlabeled.pos),
        naive.power = mean(naive.pwr)
    )


smry.otpt %>%
    ungroup() |>
    arrange(prp.nondet)

smry.otpt %>%
    ungroup() %>%
    select(-c(naive.or.mn, naive.p.mn, naive.p.sd, eff.sz)) %>%
    arrange(naive.power)

pandoc.table(smry.otpt %>% ungroup(),
    style = "rmarkdown", digits = 3, split.table = Inf
)

print(difftime(Sys.time(), now, units = "mins"))


smry.otpt %>%
    ungroup() %>%
    select(n.cases, n.naiv.control, prp.nondet, naive.power) %>%
    reframe(prp.nondet = prp.nondet, zzz = naive.power / mean(naive.power[prp.nondet == 0]))


pandoc.table(smry.otpt %>% ungroup() %>% arrange(prp.nondet),
    style = "rmarkdown", digits = 3, split.table = Inf
)
```


```{r}
# df.plot <- smry.otpt %>%
#     ungroup() %>%
#     select(n.cases, n.naiv.control, prp.nondet, naive.power) %>%
#     reframe(prp.nondet = prp.nondet, zzz = naive.power /
# mean(naive.power[prp.nondet == 0])) %>%
#     mutate(prp.nondet = prp.nondet * 100, zzz = zzz * 100)

# this plots negative power losss
df.plot <- smry.otpt %>%
    ungroup() %>%
    select(n.cases, n.naiv.control, prp.nondet, naive.power) %>%
    reframe(prp.nondet = prp.nondet, zzz = naive.power /
        mean(naive.power[prp.nondet == 0])) %>%
    mutate(prp.nondet = prp.nondet * 100, zzz = (zzz * 100) - 100)


# tiff("powerloss3.tiff", units = "in", width = 6, height = 4, res = 600)

ggplot(data = df.plot, aes(x = prp.nondet, zzz)) +
    geom_line(color = "red") +
    geom_point() +
    xlab("Percent of undetected positives in the control group") +
    ylab("Relative power loss (%)") +
    ggtitle("Power loss from positive cases in the negative control group")
# dev.off()
```


Sandbox in this chunk
```{r,include=FALSE}
n.cases <- 100
n.naiv.control <- 100
eff.sz <- .5
prp.nondet <- .05


### numbers of things
n.dogs <- n.cases + n.naiv.control # total number of dogs


# control group is made up of the unlabeled pos and the real unaffected
# number of positive dogs who are unlabled
unlab.pos <- round(n.naiv.control * prp.nondet)
n.tru.unaff <- n.naiv.control - unlab.pos # the number of unaffected

n.tru.aff <- n.cases + unlab.pos # the number of positive dogs in both groups

tru.labels <- c(rep(1, n.tru.aff), rep(0, (n.dogs - n.tru.aff)))
naive.labels <- c(rep(1, n.cases), rep(0, n.naiv.control))




### sample X
# this is where the stocastic part comes in

x.tru.affected <- rnorm(n.cases, mean = eff.sz)

x.tru.control <- c(
    rnorm(unlab.pos, mean = eff.sz),
    rnorm(n.tru.unaff)
)

naive.dat <- data.frame(
    X = as.numeric(c(x.tru.affected, x.tru.control) > 0.5),
    y = naive.labels
)

table(naive.dat$y, naive.dat$X)


cohen_w <- rcompanion::cohenW(naive.dat$y, naive.dat$X)
pwr.chisq.test(w = cohen_w, df = 1, N = 200, sig.level = 0.05)$power
```


Run three at a time
```{r}
now <- Sys.time()


parmlist <- list(
    rep(10000, 5), # interations; num.iter
    rep(100, 5), # n.cases
    rep(100, 5), # number unlabeled
    rep(.5, 5), # eff.sz for normals
    seq(from = 0, to = 0.1, length.out = 5)
) # non-detection proportion



big.otpt_100_100 <- bind_rows(pmap(parmlist, sim.fun),
    .id = "column_label"
) # unlists and binds rows

smry.otpt_100_100 <- big.otpt_100_100 %>%
    group_by(n.cases, n.naiv.control, prp.nondet, eff.sz) %>%
    dplyr::summarise(
        prp.nondet = mean(prp.nondet),
        eff.sz = mean(eff.sz),
        naive.or.mn = mean(naive.or),
        naive.p.mn = mean(naive.p),
        naive.p.sd = sd(naive.p),
        unlabeled.pos = mean(unlabeled.pos),
        naive.power = mean(naive.pwr)
    )


pandoc.table(smry.otpt_100_100 %>% ungroup() %>% arrange(prp.nondet),
    style = "rmarkdown", digits = 3, split.table = Inf
)


parmlist <- list(
    rep(10000, 5), # interations; num.iter
    rep(150, 5), # n.cases
    rep(50, 5), # number unlabeled
    rep(.5, 5), # eff.sz for normals
    seq(from = 0, to = 0.1, length.out = 5)
) # non-detection proportion




big.otpt_150_50 <- bind_rows(pmap(parmlist, sim.fun),
    .id = "column_label"
) # unlists and binds rows

smry.otpt_150_50 <- big.otpt_150_50 %>%
    group_by(n.cases, n.naiv.control, prp.nondet, eff.sz) %>%
    dplyr::summarise(
        prp.nondet = mean(prp.nondet),
        eff.sz = mean(eff.sz),
        naive.or.mn = mean(naive.or),
        naive.p.mn = mean(naive.p),
        naive.p.sd = sd(naive.p),
        unlabeled.pos = mean(unlabeled.pos),
        naive.power = mean(naive.pwr)
    )


pandoc.table(smry.otpt_150_50 %>% ungroup() %>% arrange(prp.nondet),
    style = "rmarkdown", digits = 3, split.table = Inf
)


parmlist <- list(
    rep(10000, 5), # interations; num.iter
    rep(50, 5), # n.cases
    rep(150, 5), # number unlabeled
    rep(.5, 5), # eff.sz for normals
    seq(from = 0, to = 0.1, length.out = 5)
) # non-detection proportion



big.otpt_50_150 <- bind_rows(pmap(parmlist, sim.fun),
    .id = "column_label"
) # unlists and binds rows



smry.otpt_50_150 <- big.otpt_50_150 %>%
    group_by(n.cases, n.naiv.control, prp.nondet, eff.sz) %>%
    dplyr::summarise(
        prp.nondet = mean(prp.nondet),
        eff.sz = mean(eff.sz),
        naive.or.mn = mean(naive.or),
        naive.p.mn = mean(naive.p),
        naive.p.sd = sd(naive.p),
        unlabeled.pos = mean(unlabeled.pos),
        naive.power = mean(naive.pwr)
    )



pandoc.table(smry.otpt_50_150 %>% ungroup() %>% arrange(prp.nondet),
    style = "rmarkdown", digits = 3, split.table = Inf
)

difftime(Sys.time(), now)
```

plot all three curves
```{r}
# df.plot <- smry.otpt %>%
#     ungroup() %>%
#     select(n.cases, n.naiv.control, prp.nondet, naive.power) %>%
#     reframe(prp.nondet = prp.nondet, zzz = naive.power /
# mean(naive.power[prp.nondet == 0])) %>%
#     mutate(prp.nondet = prp.nondet * 100, zzz = zzz * 100)

# this plots negative power losss
df.plot1 <- smry.otpt_100_100 %>%
    ungroup() %>%
    select(prp.nondet, naive.power) %>%
    mutate(aaa = "1", zzz = naive.power)

df.plot2 <- smry.otpt_150_50 %>%
    ungroup() %>%
    select(prp.nondet, naive.power) %>%
    mutate(aaa = "2", zzz = naive.power)


df.plot3 <- smry.otpt_50_150 %>%
    ungroup() %>%
    select(prp.nondet, naive.power) %>%
    mutate(aaa = "3", zzz = naive.power)

df.plot <- bind_rows(df.plot1, df.plot2, df.plot3)

# tiff("powerloss_all_three.tiff", units = "in", width = 6, height = 4, res = 600

ggplot(data = df.plot, aes(x = prp.nondet, zzz, color = aaa)) +
    geom_line()
# dev.off()
```

plot all three curves this just changes the scale from the
above plot
```{r}
# df.plot <- smry.otpt %>%
#     ungroup() %>%
#     select(n.cases, n.naiv.control, prp.nondet, naive.power) %>%
#     reframe(prp.nondet = prp.nondet, zzz = naive.power /
# mean(naive.power[prp.nondet == 0])) %>%
#     mutate(prp.nondet = prp.nondet * 100, zzz = zzz * 100)

# this plots negative power losss
df.plot1 <- smry.otpt_100_100 %>%
    ungroup() %>%
    select(prp.nondet, naive.power) %>%
    reframe(prp.nondet = prp.nondet, zzz = naive.power) %>%
    mutate(aaa = "1", prp.nondet = prp.nondet * 100, zzz = zzz * 100)

df.plot2 <- smry.otpt_150_50 %>%
    ungroup() %>%
    select(prp.nondet, naive.power) %>%
    reframe(prp.nondet = prp.nondet, zzz = naive.power) %>%
    mutate(aaa = "2", prp.nondet = prp.nondet * 100, zzz = zzz * 100)


df.plot3 <- smry.otpt_50_150 %>%
    ungroup() %>%
    select(prp.nondet, naive.power) %>%
    reframe(prp.nondet = prp.nondet, zzz = naive.power) |>
    mutate(aaa = "3", prp.nondet = prp.nondet * 100, zzz = zzz * 100)

df.plot <- bind_rows(df.plot1, df.plot2, df.plot3)

# tiff("powerloss_all_three.tiff", units = "in", width = 6, height = 4, res = 600

ggplot(data = df.plot, aes(x = prp.nondet, zzz, color = aaa)) +
    geom_line()
# dev.off()
```

The plot below is power loss normlize to the best power, which is 
balance groups (100,100) and 0 prp.nondet.
```{r, echo=FALSE}
df.plot1 <- smry.otpt_100_100 %>%
    ungroup() %>%
    select(n.cases, n.naiv.control, prp.nondet, naive.power) %>%
    reframe(prp.nondet = prp.nondet, zzz = naive.power /
        mean(naive.power[prp.nondet == 0])) %>%
    mutate(aaa = "1", prp.nondet = prp.nondet * 100, zzz = (zzz * 100) - 100)

mn_100_100 <- as.numeric(smry.otpt_100_100 %>%
    ungroup() %>%
    select(prp.nondet, naive.power) %>%
    reframe(mean(naive.power[prp.nondet == 0])))

df.plot2 <- smry.otpt_150_50 %>%
    ungroup() %>%
    select(n.cases, n.naiv.control, prp.nondet, naive.power) %>%
    reframe(prp.nondet = prp.nondet, zzz = naive.power / mn_100_100) %>%
    mutate(aaa = "2", prp.nondet = prp.nondet * 100, zzz = (zzz * 100) - 100)

df.plot3 <- smry.otpt_50_150 %>%
    ungroup() %>%
    select(n.cases, n.naiv.control, prp.nondet, naive.power) %>%
    reframe(prp.nondet = prp.nondet, zzz = naive.power / mn_100_100) %>%
    mutate(aaa = "3", prp.nondet = prp.nondet * 100, zzz = (zzz * 100) - 100)

df.plot <- bind_rows(df.plot1, df.plot2, df.plot3)

tiff("powerloss_all_three.tiff", units = "in", width = 7, height = 4, res = 600)

ggplot(data = df.plot, aes(x = prp.nondet, zzz, color = aaa)) +
    geom_line() +
    xlab("Percent undetected postitives in the control group") +
    ylab("Percent power loss") +
    labs(
        title = "Percent power loss from postitive cases in the GWAS control group",
        subtitle = "Relative to max power: 100 cases, 100 controls with no undetected positives",
        caption = "Effect size = 0.5, alpha = 0.05",
        color = "Naive sample sizes"
    ) +
    theme(
        plot.title = element_text(size = 14), # Center title position and size
        plot.subtitle = element_text(size = 10), # Center subtitle
        plot.caption = element_text(face = "italic") # move caption to the left
    )+scale_color_discrete(labels=c('100 cases, 100 controls', '150 cases, 50 controls', '50 cases, 150 controls'))


dev.off()
```