---
title: "sim"
author: Report by Rich Evans
date: 2023
output:
    html_document


---

This simulation implements part of:

Gu W, Swihart RK. Absent or undetected? Effects of non-detection of species occurrence on wildlife–habitat models. Biological conservation. 2004 Apr 1;116(2):195-203.

But they are random sampling a wildlife area. They randomly pick occupied areas in the simulation. In other words they have a grid and randomly assign the occupied areas. 

I have retrospective data there. I use a mixture of normals because I have sample sizes I need to investigate

```{r,include=FALSE}
#install.packages("pacman")
pacman::p_load(
       data.table, # to group and clean data
       tidyverse,  # allows use of pipe (%>%) function in this chapter
       janitor,     #for clean_names and tabyl
       broom,   #tidy output, use tidy(xxx)
       gtsummary,  #nice tables
       magrittr, # for %<>%
       effectsize,
       pROC,
       rstatix,
       pwr,   # power calculations
       pander, #to write tables as simple text for cut-and-paste
       knitr,
        kableExtra
   ) 
```




```{r,include=FALSE,warning=FALSE,message=FALSE}


n.cases = 100
n.naiv.control = 100
prp.nondet = .4

out1 <- NULL
out2 <- NULL
out3 <- NULL
out4 <- NULL

n.dogs<-n.cases + n.naiv.control

unlab.pos <- round(n.naiv.control*prp.nondet)
n.tru.aff<-n.cases +unlab.pos

tru.labels<- c(rep(1,n.tru.aff),rep(0,(n.dogs-n.tru.aff)))
naive.labels<- c(rep(1,n.cases),rep(0,n.naiv.control))


  
###sample X
# this is where the stocastic part comes in
  
# x.tru.affected <- sort(rnorm(n.tru.aff,mean=eff.sz),decreasing=FALSE)
# x.tru.control<-sort(rnorm(n.dogs-n.tru.aff),decreasing=TRUE)
####NEED A MIXTURE FOR THE UNLABELED
x.tru.affected <- runif(n.tru.aff,.5,1)
x.tru.control<-c(runif(unlab.pos,.5,1),
                 runif(n.dogs-n.tru.aff-unlab.pos,0,.5))
### estimate coeff naive model



naive.dat<-data.frame(X=c(x.tru.affected,x.tru.control),
                       y=naive.labels)

 hist(naive.dat$X)
 
 
m.naive<-glm(y~X,data=naive.dat, family = binomial())

m.naive.pred <-predict(m.naive, type = "link")

as.numeric(roc(tru.labels ~ m.naive.pred)$auc)


out1 <- c(out1,as.numeric(roc(tru.labels ~ m.naive.pred)$auc))
out2<- c(out2,sum(tru.labels))
out3<- c(out3,sum(naive.labels))
 

                    
```


