---
title: "sim"
author: Report by Rich Evans
date: 2023
output:
    html_document


---

This simulation implements part of:

Gu W, Swihart RK. Absent or undetected? Effects of non-detection of species occurrence on wildlife–habitat models. Biological conservation. 2004 Apr 1;116(2):195-203.

But they are random sampling a wildlife area. They randomly pick occupied areas in the simulation. In other words they have a grid and randomly assign the occupied areas. 

I have retrospective data there. I use a mixture of normals because I have sample sizes I need to investigate

```{r,include=FALSE}
# install.packages("pacman")
pacman::p_load(
    data.table, # to group and clean data
    tidyverse, # allows use of pipe (%>%) function in this chapter
    janitor, # for clean_names and tabyl
    broom, # tidy output, use tidy(xxx)
    gtsummary, # nice tables
    magrittr, # for %<>%
    effectsize,
    pROC,
    rstatix,
    pwr, # power calculations
    pander, # to write tables as simple text for cut-and-paste
    knitr,
    rcompanion, # for cohen's w
    kableExtra,
    bindata,
    here
)

here::i_am("README.md")
```




```{r,include=FALSE}
sim.fun <- function(
    num.iter = NULL,
    n.cases = NULL,
    n.naiv.control = NULL,
    eff.sz = NULL,
    prp.nondet = NULL, ...) {
    out1 <- NULL
    out2 <- NULL
    out3 <- NULL
    out4 <- NULL

    ### numbers of things
    n.dogs <- n.cases + n.naiv.control # total number of dogs


    # control group is made up of the unlabeled pos and the real unaffected
    # number of positive dogs who are unlabled
    unlab.pos <- round(n.naiv.control * prp.nondet)
    n.tru.unaff <- n.naiv.control - unlab.pos # the number of unaffected

    n.tru.aff <- n.cases + unlab.pos # the number of positive dogs in both groups

    tru.labels <- c(rep(1, n.tru.aff), rep(0, (n.dogs - n.tru.aff)))
    naive.labels <- c(rep(1, n.cases), rep(0, n.naiv.control))



    for (i in 1:num.iter) {
        ### sample X
        # this is where the stocastic part comes in

        x.tru.affected <- rnorm(n.cases, mean = eff.sz)

        x.tru.control <- c(
            rnorm(unlab.pos, mean = eff.sz),
            rnorm(n.tru.unaff)
        )

        naive_X <- c(x.tru.affected, x.tru.control)
        naive_X_z <- (naive_X - mean(naive_X)) / sd(naive_X)

        # naive.dat <- data.frame(
        #     X = as.numeric(pnorm(naive_X_z) > runif(200, min = -.1, max = .1)),
        #     y = naive.labels
        # )

        naive.dat <- data.frame(
            X = as.numeric(naive_X_z > 0),
            y = naive.labels
        )

        m.naive <- glm(y ~ X, data = naive.dat, family = binomial())

        naive.coeff <- m.naive$coefficients["X"]
        naive.p <- summary(m.naive)$coef[2, 4] # pvalue

        out1 <- c(out1, naive.p)
        out2 <- c(out2, exp(naive.coeff))
        out3 <- c(out3, unlab.pos)

        # temp.out1 <- rcompanion::cohenW(naive.dat$y, naive.dat$X) # cohen_w


        # temp.out4 <- pwr.chisq.test(w = temp.out1, df = 1, N = 200, sig.level = 0.05)$power
        # out4 <- c(out4, temp.out4)

temp.out1<-cohens_d(naive_X_z ~ naive.dat$y)[[1]]


temp.out4<-pwr.2p2n.test(temp.out1,n.cases,n.naiv.control,0.05)[[5]]
out4 <- c(out4, temp.out4)

    } # loop

    return(list(
        n.cases = n.cases,
        n.naiv.control = n.naiv.control,
        prp.nondet = prp.nondet,
        eff.sz = eff.sz,
        naive.p = out1,
        naive.or = out2,
        unlabeled.pos = out3,
        naive.pwr = out4
    ))
}
```


this version samples from a binomial
```{r,include=FALSE}
sim.fun <- function(
    num.iter = NULL,
    n.cases = NULL,
    n.naiv.control = NULL,
    eff.sz = NULL,
    prp.nondet = NULL, ...) {
    out1 <- NULL
    out2 <- NULL
    out3 <- NULL
    out4 <- NULL

    ### numbers of things
    n.dogs <- n.cases + n.naiv.control # total number of dogs


    # control group is made up of the unlabeled pos and the real unaffected
    # number of positive dogs who are unlabled
    unlab.pos <- round(n.naiv.control * prp.nondet)
    n.tru.unaff <- n.naiv.control - unlab.pos # the number of unaffected

    n.tru.aff <- n.cases + unlab.pos # the number of positive dogs in both groups

    tru.labels <- c(rep(1, n.tru.aff), rep(0, (n.dogs - n.tru.aff)))
    naive.labels <- c(rep(1, n.cases), rep(0, n.naiv.control))



    for (i in 1:num.iter) {
        ### sample X
        # this is where the stocastic part comes in

        x.real.affected <- rbinom(n.cases, 1, eff.sz + .2)

# eff.sz is the probability above baseline of being affected without the snp
# the binomials determine who is affected.
# Pan would have determined "prob" from a logistic formula
# where the covariate is the snp and the snp comes from a latent MVN.
# For Pan, it would be sample a MVN, use a cutoff to determine mutation, then
# use mutation (0,1,2) with the OR parameter and baseline rate to calc "prob"
# in the usual formula 1/1+exp(-baseline-or*snp))

        x.real.control <- c(
            rbinom(unlab.pos, 1, prob = eff.sz + .2),   #0.2 is the background disease prevalence
            rbinom(n.tru.unaff, 1, prob = 0.2)
        )

        naive_X <- c(x.real.affected, x.real.control)
    

    
        naive.dat <- data.frame(
            X = naive_X,
            y = naive.labels
        )

         temp.out1 <- as.numeric(rcompanion::cohenW(naive.dat$y, naive.dat$X)) # cohen_w
        out1 <- c(out1, temp.out1)

         temp.out4 <- pwr.chisq.test(w = temp.out1, df = 1, N = 200, sig.level = 0.05)$power
         out4 <- c(out4, temp.out4)

        junk<-table(naive.dat$y, naive.dat$X)
        out2 <- c(out2, (junk[1,1]*junk[2,2])/(junk[2,1]*junk[1,2]))
    } # loop

    return(list(
        n.cases = n.cases,
        n.naiv.control = n.naiv.control,
        prp.nondet = prp.nondet,
        eff.sz = eff.sz,
        naive.pwr = out4,
        cohen = out1,
        or = out2
    ))
}
```

```{r, echo=FALSE}
rcompanion::cohenW(c(rep(1,50),rep(0,100)), c(rep(1,60),rep(0,90)))

rstatix::chisq_test(c(rep(1,50),rep(0,100)), c(rep(1,60),rep(0,90)))

rcompanion::cohenW(c(rep(1,100),rep(0,50)), c(rep(1,105),rep(0,45)))
rstatix::chisq_test(c(rep(1,100),rep(0,50)), c(rep(1,105),rep(0,45)))


```

Run three at a time-makes all the data for the plot
```{r}
now <- Sys.time()


parmlist <- list(
    rep(5001, 3), # interations; num.iter
    rep(100, 3), # n.cases
    rep(100, 3), # number unlabeled
    rep(.21, 3), # eff.sz for normals
    seq(from = 0, to = 0.1, length.out = 3)
) # non-detection proportion

big.otpt_100_100 <- bind_rows(pmap(parmlist, sim.fun),
    .id = "column_label"
) # unlists and binds rows

smry.otpt_100_100 <- big.otpt_100_100 %>%
    group_by(n.cases, n.naiv.control, prp.nondet, eff.sz) %>%
    dplyr::summarise(
        prp.nondet = mean(prp.nondet),
        eff.sz = mean(eff.sz),
        naive.power = mean(naive.pwr),
        cohen = mean(cohen),
        or = mean(or)
    )

pandoc.table(smry.otpt_100_100 %>% ungroup() %>% arrange(prp.nondet),
    style = "rmarkdown", digits = 3, split.table = Inf
)


parmlist <- list(
    rep(5001, 3), # interations; num.iter
    rep(150, 3), # n.cases
    rep(50, 3), # number unlabeled
    rep(.21, 3), # eff.sz for normals
    seq(from = 0, to = 0.1, length.out = 3)
) # non-detection proportion




big.otpt_150_50 <- bind_rows(pmap(parmlist, sim.fun),
    .id = "column_label"
) # unlists and binds rows

smry.otpt_150_50 <- big.otpt_150_50 %>%
    group_by(n.cases, n.naiv.control, prp.nondet, eff.sz) %>%
    dplyr::summarise(
        prp.nondet = mean(prp.nondet),
        eff.sz = mean(eff.sz),
        naive.power = mean(naive.pwr),
        cohen = mean(cohen),
        or = mean(or)
    )


pandoc.table(smry.otpt_150_50 %>% ungroup() %>% arrange(prp.nondet),
    style = "rmarkdown", digits = 3, split.table = Inf
)


parmlist <- list(
    rep(5001, 3), # interations; num.iter
    rep(50, 3), # n.cases
    rep(150, 3), # number unlabeled
    rep(.21, 3), # eff.sz for normals
    seq(from = 0, to = 0.1, length.out = 3)
) # non-detection proportion



big.otpt_50_150 <- bind_rows(pmap(parmlist, sim.fun),
    .id = "column_label"
) # unlists and binds rows



smry.otpt_50_150 <- big.otpt_50_150 %>%
    group_by(n.cases, n.naiv.control, prp.nondet, eff.sz) %>%
    dplyr::summarise(
        prp.nondet = mean(prp.nondet),
        eff.sz = mean(eff.sz),
        naive.power = mean(naive.pwr),
        cohen = mean(cohen),
        or = mean(or)
    )



pandoc.table(smry.otpt_50_150 %>% ungroup() %>% arrange(prp.nondet),
    style = "rmarkdown", digits = 3, split.table = Inf
)

difftime(Sys.time(), now)
```


usually use this one
The plot below is power loss normlize to the best power, which is 
balance groups (100,100) and 0 prp.nondet.
```{r, echo=FALSE}
df.plot1 <- smry.otpt_100_100 %>%
    ungroup() %>%
    select(n.cases, n.naiv.control, prp.nondet, naive.power) %>%
    reframe(prp.nondet = prp.nondet, zzz = naive.power /
        mean(naive.power[prp.nondet == 0])) %>%
    mutate(aaa = "1", prp.nondet = prp.nondet * 100, zzz = (zzz * 100) - 100)


mn_100_100 <- as.numeric(smry.otpt_100_100 %>%
    ungroup() %>%
    select(prp.nondet, naive.power) %>%
    reframe(mean(naive.power[prp.nondet == 0])))

df.plot2 <- smry.otpt_150_50 %>%
    ungroup() %>%
    select(n.cases, n.naiv.control, prp.nondet, naive.power) %>%
    reframe(prp.nondet = prp.nondet, zzz = naive.power / mn_100_100) %>%
    mutate(aaa = "2", prp.nondet = prp.nondet * 100, zzz = (zzz * 100) - 100)

df.plot3 <- smry.otpt_50_150 %>%
    ungroup() %>%
    select(n.cases, n.naiv.control, prp.nondet, naive.power) %>%
    reframe(prp.nondet = prp.nondet, zzz = naive.power / mn_100_100) %>%
    mutate(aaa = "3", prp.nondet = prp.nondet * 100, zzz = (zzz * 100) - 100)

df.plot <- bind_rows(df.plot1, df.plot2, df.plot3)

tiff("powerloss_all_three_80percent.tiff", units = "in", width = 7, height = 4, res = 600)

ggplot(data = df.plot, aes(x = prp.nondet, zzz, color = aaa)) +
    geom_smooth() +
    xlab("Percent undetected postitives in the control group") +
    ylab("Percent power loss") +
    labs(
        title = "Percent power loss from positive cases in the GWAS control group",
        subtitle = "Relative to reference power: 100 cases, 100 controls with no undetected positives",
        caption = "Reference power = 0.82, alpha = 0.05",
        color = "Naive sample sizes"
    ) +
    theme(
        plot.title = element_text(size = 14), # Center title position and size
        plot.subtitle = element_text(size = 10), # Center subtitle
        plot.caption = element_text(face = "italic") # move caption to the left
    ) +
    scale_color_discrete(labels = c("100 cases, 100 controls", "150 cases, 50 controls", "50 cases, 150 controls"))


dev.off()
```

# Sandbox

```{r,include=FALSE}


    n.cases <- 150
    n.naiv.control <- 50
    eff.sz <- 0.5
    prp.nondet <- 0

    ### numbers of things
    n.dogs <- n.cases + n.naiv.control # total number of dogs


    # control group is made up of the unlabeled pos and the real unaffected
    # number of positive dogs who are unlabled
    unlab.pos <- round(n.naiv.control * prp.nondet)
    n.tru.unaff <- n.naiv.control - unlab.pos # the number of unaffected

    n.tru.aff <- n.cases + unlab.pos # the number of positive dogs in both groups

    tru.labels <- c(rep(1, n.tru.aff), rep(0, (n.dogs - n.tru.aff)))
    naive.labels <- c(rep(1, n.cases), rep(0, n.naiv.control))

table(tru.labels)
table(naive.labels)


        ### sample X
        # this is where the stocastic part comes in

        x.tru.affected <- rnorm(n.cases, mean = eff.sz)

        x.tru.control <- c(
            rnorm(unlab.pos, mean = eff.sz),
            rnorm(n.tru.unaff)
        )

hist(x.tru.affected)
hist(x.tru.control)

        naive_X <- c(x.tru.affected, x.tru.control)
        naive_X_z <- (naive_X - mean(naive_X)) / sd(naive_X)

        # naive.dat <- data.frame(
        #     X = as.numeric(pnorm(naive_X_z) > runif(200, min = -.01, max = .01)),
        #     y = naive.labels
        # )

                naive.dat <- data.frame(
            X = as.numeric(naive_X_z > runif(200, min = -.01, max = .01)),
            y = naive.labels
        )
hist(naive_X_z)

temp.out1<-cohens_d(naive_X_z ~ naive.dat$y)[[1]]
#temp.out1<-rcompanion::cohenW(naive.dat$y, naive.dat$X) # cohen_w

temp.out1
#pwr.chisq.test(w = temp.out1, df = 1, N = 200, sig.level = 0.05)$power

pwr.2p2n.test(temp.out1,n.cases,n.naiv.control,0.05)[[5]]
```


http://www.biostat.umn.edu/~weip/prog.html
http://www.biostat.umn.edu/~weip/prog/BasuPanGE11/simRareSNP.R

```{r, echo=FALSE}
########simulate SNPs from a latent multivariate Gaussian variable with
######## a CS(rho) correlation structure. 

########Input: 
#               ORs: association in OR between the causal SNPs and outcome;
#                   implicitly,  # of causal SNPs = length(ORs);
#               n0: # of controls;
#               n1: # of cases ;
#               nSNP: # of noise/marker SNPs to be generated;
#               rho: the parameter in CS(rho) for the latent multivariate
#                    Gaussian variable to be discretized to SNPs;
#                    rho=0 means all SNPs are independent;
#               MAF0slow, MAF0sup: MAFs for the causal SNPs from 
#                                   Unif(MAF0slow, MAF0sup);
#               MAFslow, MAFsup: MAF's of the noise SNPs are drawn from
#                                 Unif(MAFslow, MAFsup);
#               p0: background disease prevalence; i.e. true logistic reg model's intercept=log(p0/(1-p0))
########Output: a list of the binary outcome Y (=0 or 1) and SNPs (=0, 1 or 2);
#               Y is a vector of length n=n1+n0; X is a matrix of n by nSNP;
#               and location indices for causal SNPs.

simRareSNP<-function(ORs, n0=500, n1=500, nSNP=0, rho=0,
                 MAF0slow=0.001, MAF0sup=0.01,
                 MAFslow=0.001, MAFsup=0.01, p0=0.05){

nSNP0<-length(ORs)
q<-nSNP+nSNP0

###############index for causal SNPs; mix causal and noise SNPs:
############### SNP j is causal iff SNP0index[j]=1 or -1; =0 o/w
signORs<-rep(1, nSNP0)
signORs[ORs<1]<-0-1
SNP0indx<-sample(c(signORs,  rep(0, nSNP)), q, replace = FALSE)

R<-matrix(1, nrow=q, ncol=q)
for(i in 1:q)
  for(j in 1:q)
    if (i!=j) R[i, j]<-rho
svd.R<-svd(R)
R1<-svd.R$u %*% diag(sqrt(svd.R$d))

##background disease prev = p0
b0<-log(p0/(1-p0))
logORs<-log(ORs)

MAF0s<-runif(nSNP0, MAF0slow, MAF0sup)
MAFs1<-runif(nSNP, MAFslow, MAFsup)
cutoff0<-qnorm(MAF0s)
cutoff1<-qnorm(MAFs1)

cutoff<-rep(0, q)
cutoff[SNP0indx!=0]=cutoff0
cutoff[SNP0indx==0]=cutoff1

X<-matrix(0, nrow=n0+n1, ncol=q)
Y<-rep(0, n0+n1); Y[(n0+1):(n0+n1)]<-1
i<-1
#sampling controls:
while ( i <= n0){
  X0<-rnorm(q, 0, 1) #: X0 ~ MVN(0, I)
  X1<-R1 %*% X0   #: X1 ~ MVN(0, R)
  X2<-ifelse(X1<cutoff, 1, 0)
  X0<-rnorm(q, 0, 1) #: X0 ~ MVN(0, I)
  X1<-R1 %*% X0   #: X1 ~ MVN(0, R)
  X3<-ifelse(X1<cutoff, 1, 0)
  X4<-X2+ X3
  pr<-1/(1 + exp(-(b0 + sum(logORs * X4[SNP0indx!=0]))))
  Y1<-sample(c(0, 1), 1, prob=c(1-pr, pr))
  if (Y1==0){
    X[i, ]<-X4
    i<-i+1
    }
  }
#sampling cases:
while ( i <= n0+n1){
  X0<-rnorm(q, 0, 1) #: X0 ~ MVN(0, I)
  X1<-R1 %*% X0   #: X1 ~ MVN(0, R)
  X2<-ifelse(X1<cutoff, 1, 0)
  X0<-rnorm(q, 0, 1) #: X0 ~ MVN(0, I)
  X1<-R1 %*% X0   #: X1 ~ MVN(0, R)
  X3<-ifelse(X1<cutoff, 1, 0)
  X4<-X2+ X3
  pr<-1/(1 + exp(-(b0 + sum(logORs * X4[SNP0indx!=0]))))
  Y1<-sample(c(0, 1), 1, prob=c(1-pr, pr))
  if (Y1==1){
    X[i, ]<-X4
    i<-i+1
    }
  }

list(Y=Y, X=X, SNP0indx=SNP0indx)

}
```

```{r, echo=FALSE}
junk<-simRareSNP(5,n0=100,n1=100, p=.2)

str(junk)
table(junk$Y,junk$X)
```